---
title: "Epic Cosmos mental health ED visits"
subtitle: "`epic_mental_health` standardized output"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    fig_width: 8
    fig_height: 4.6
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dpi = 140, fig.retina = 2, out.width = "100%"
)
```

```{r libs, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(sf)
```

```{r palette, include = FALSE}
# Palette: slots 1-3 of the validated categorical set, plus neutral ink. Magnitude
# is carried by position everywhere except the choropleth, which uses the
# single-hue blue sequential ramp.
blue   <- "#2a78d6"
orange <- "#eb6834"
aqua   <- "#1baf7a"
ink    <- "#0b0b0b"
ink2   <- "#52514e"
muted  <- "#898781"
grid_c <- "#e1e0d9"
axis_c <- "#c3c2b7"
surf   <- "#fcfcfb"

seq_blue <- c("#cde2fb", "#9ec5f4", "#6da7ec", "#3987e5",
              "#256abf", "#184f95", "#0d366b")

theme_pop <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background   = element_rect(fill = surf, colour = NA),
      panel.background  = element_rect(fill = surf, colour = NA),
      panel.grid.major  = element_line(colour = grid_c, linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      axis.line.x       = element_line(colour = axis_c, linewidth = 0.4),
      axis.ticks        = element_blank(),
      axis.text         = element_text(colour = muted, size = rel(0.85)),
      axis.title        = element_text(colour = ink2, size = rel(0.9)),
      plot.title        = element_text(colour = ink, face = "bold", size = rel(1.05)),
      plot.subtitle     = element_text(colour = ink2, size = rel(0.9)),
      plot.caption      = element_text(colour = muted, size = rel(0.8), hjust = 0),
      strip.text        = element_text(colour = ink, face = "bold",
                                       size = rel(0.85), hjust = 0),
      legend.position   = "top",
      legend.justification = "left",
      legend.title      = element_blank(),
      legend.text       = element_text(colour = ink2, size = rel(0.85)),
      plot.margin       = margin(10, 14, 10, 10)
    )
}
theme_set(theme_pop())

theme_map <- function() {
  theme_pop() +
    theme(
      panel.grid.major = element_blank(),
      axis.line.x      = element_blank(),
      axis.text        = element_blank(),
      axis.title       = element_blank(),
      legend.position  = "right",
      legend.title     = element_text(colour = ink2, size = rel(0.85))
    )
}
```

```{r load, include = FALSE}
raw <- vroom::vroom(
  "standard/data.csv.gz",
  col_types = vroom::cols(geography = "c", time = "c", sex = "c", .default = "d")
)

dx_lab <- c(
  suicidal_behavior = "Suicidal behavior",
  mood              = "Mood",
  substance_use     = "Substance use",
  psychosis         = "Psychosis",
  behavioral        = "Behavioral",
  eating_disorders  = "Eating disorders",
  other             = "Other MH",
  all_cause         = "All ED encounters"
)

d_all <- raw %>%
  pivot_longer(starts_with("epic_"), names_to = "col", values_to = "v") %>%
  mutate(
    kind = if_else(grepl("_suppressed_flag$", col), "flag", "value"),
    base = sub("_suppressed_flag$", "", col)
  ) %>%
  select(-col) %>%
  pivot_wider(names_from = kind, values_from = v) %>%
  mutate(
    stat = if_else(
      grepl("^epic_ed_pct_visits_", base),
      "pct_visits",
      sub("^epic_ed_los_([a-z]+)_.*$", "\\1", base)
    ),
    dx = if_else(
      grepl("^epic_ed_pct_visits_", base),
      sub("^epic_ed_pct_visits_", "", base),
      sub("^epic_ed_los_[a-z]+_", "", base)
    ),
    dx    = factor(dx, levels = names(dx_lab), labels = dx_lab),
    date  = as.Date(time),
    is_mh = dx != "All ED encounters"
  ) %>%
  select(geography, date, sex, dx, stat, value, flag, is_mh)

# -- Suppressed cells are excluded from every analysis in this document. --------
# A flag of 1 means the source withheld the cell (10 or fewer encounters). The
# ingest drops nothing, so a flag of 1 and a missing value coincide exactly and
# this one filter is the whole exclusion rule.
#
# Negative durations are NOT excluded: they are what the source published and are
# retained here as such. They are confined to the min statistic (~17% of cells)
# and four avg cells, none of them national, so no figure below is materially
# affected - but the negatives are counted in the note that follows so the
# decision is visible rather than implicit.
n_before <- nrow(d_all)
d <- d_all %>% filter(flag == 0, !is.na(value))
excluded <- n_before - nrow(d)
n_neg <- sum(d$value < 0)

all_fips <- vroom::vroom("../../resources/all_fips.csv.gz", show_col_types = FALSE) %>%
  filter(nchar(geography) == 2) %>%
  select(geography, geography_name)

d <- left_join(d, all_fips, by = "geography")

win_start <- max(d$date) - 365
```

```{r pool-mh, include = FALSE}
# -- "All mental health visits" -------------------------------------------------
# The source publishes no encounter counts, so the seven diagnosis buckets cannot
# be pooled directly. But epic_ed_pct_visits_* is each bucket's share of the same
# denominator (all ED encounters in that geography/month/sex), so it is
# proportional to the bucket's encounter count and can serve as the weight.
#
#   pooled statistic = sum(w_i * stat_i) / sum(w_i),   w_i = pct_visits_i
#
# This is exact for the mean. For the median it is an approximation - a weighted
# average of bucket medians is not the median of the pooled distribution - and is
# labelled as such wherever it appears.
#
# Pooling runs over observed buckets only, and a cell is emitted only if the four
# well-observed diagnoses are all present. Those four carry ~98% of national
# mental health volume, so the estimate is stable even where the sparse buckets
# (eating disorders, behavioral, other) are suppressed and drop out.
CORE <- c("Suicidal behavior", "Mood", "Substance use", "Psychosis")

wts <- d %>%
  filter(stat == "pct_visits") %>%
  select(geography, date, sex, dx, w = value)

mh_pool <- d %>%
  filter(is_mh, stat %in% c("avg", "median", "sd")) %>%
  inner_join(wts, by = c("geography", "date", "sex", "dx")) %>%
  group_by(geography, geography_name, date, sex, stat) %>%
  filter(all(CORE %in% as.character(dx))) %>%
  summarize(value = sum(value * w) / sum(w), n_dx = n(), .groups = "drop") %>%
  mutate(dx = "All mental health", is_mh = TRUE)

# Analysis frame: the seven diagnosis buckets plus the pooled mental health total
lev <- c(setdiff(unname(dx_lab), "All ED encounters"), "All mental health")

mh <- bind_rows(
  d %>% filter(is_mh, stat %in% c("avg", "median", "sd")) %>%
    select(geography, geography_name, date, sex, dx, stat, value),
  mh_pool %>% select(geography, geography_name, date, sex, dx, stat, value)
) %>%
  mutate(dx = factor(as.character(dx), levels = lev))
```

**Suppressed cells are excluded from every figure and table below.**
`r format(excluded, big.mark = ",")` of `r format(n_before, big.mark = ",")` measure
cells (`r sprintf("%.0f%%", 100 * excluded / n_before)`) are suppressed.

Some values are negative in Epic Cosmos: 9,513 in min, 4 in avg


# 1. Length of stays

```{r fig-dist, fig.height = 6.6}
# Each point is the unweighted mean of that year's monthly values. Months cannot
# be volume-weighted: the source publishes no encounter counts.
dist <- mh %>%
  filter(sex == "Overall", geography == "00", stat %in% c("median", "avg"),
         dx != "All mental health") %>%
  mutate(year = as.integer(format(date, "%Y"))) %>%
  group_by(dx, stat, year) %>%
  summarize(value = mean(value), .groups = "drop") %>%
  mutate(stat = factor(stat, c("median", "avg"), c("Median", "Mean")))

# One y order for every panel, from the all-year median, so panels are comparable
ord <- dist %>%
  filter(stat == "Median") %>%
  group_by(dx) %>%
  summarize(v = mean(value), .groups = "drop") %>%
  arrange(v) %>%
  pull(dx)

yr_lab <- c("2022" = "2022 (Jul-Dec)", "2026" = "2026 (Jan-Jun)")
dist <- dist %>%
  mutate(year_lab = ifelse(as.character(year) %in% names(yr_lab),
                           yr_lab[as.character(year)], as.character(year)))

ggplot(dist, aes(x = value, y = factor(dx, levels = ord), colour = stat)) +
  geom_point(size = 2.4) +
  facet_wrap(~ year_lab, ncol = 3) +
  scale_colour_manual(values = c(Median = blue, Mean = orange)) +
  scale_x_continuous(expand = expansion(mult = c(0.08, 0.12)),
                     breaks = scales::breaks_pretty(4)) +
  labs(
    title = "Mean and median ED stay by mental health diagnosis and year",
    subtitle = "National, all sexes. Each point averages that year's monthly values",
    x = "Minutes from ED arrival to departure", y = NULL
  )
```

# 2. Trends over time

```{r fig-trend, fig.height = 5.4}
trend <- mh %>% filter(sex == "Overall", geography == "00", stat == "median")

ggplot(trend, aes(date, value)) +
  geom_line(colour = blue, linewidth = 0.7) +
  facet_wrap(~ dx, ncol = 4, scales = "free_y") +
  scale_x_date(date_labels = "'%y", date_breaks = "1 year") +
  labs(
    title = "Median stay by diagnosis",
    x = NULL, y = "Median minutes"  )
```


## Stratified by sex

```{r fig-trend-sex, fig.height = 5.4}
trend_sex <- mh %>%
  filter(geography == "00", stat == "median", sex %in% c("Female", "Male"))

ggplot(trend_sex, aes(date, value, colour = sex)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ dx, ncol = 4, scales = "free_y") +
  scale_colour_manual(values = c(Female = blue, Male = orange)) +
  scale_x_date(date_labels = "'%y", date_breaks = "1 year") +
  labs(
    title = "Median stay by diagnosis and sex",
    subtitle = "Ambiguous stratum is omitted, it is heavily suppressed",
    x = NULL, y = "Median minutes",
  )
```


## Stratified by age



# 3. Median stay by state and diagnosis

```{r fig-map-dx, fig.height = 7.5}
# Puerto Rico ships with the usmap polygons but never appears in this source, so it
# is dropped rather than drawn as an unexplained grey shape.
us <- usmap::us_map(regions = "states") %>% filter(fips != "72")

yr_lab <- c("2022" = "2022 (Jul-Dec)", "2026" = "2026 (Jan-Jun)")

map_dx <- d %>%
  filter(is_mh, stat == "median", sex == "Overall", geography != "00") %>%
  mutate(year = as.integer(format(date, "%Y")), dx = as.character(dx)) %>%
  group_by(dx, geography, geography_name, year) %>%
  summarize(median_min = mean(value), .groups = "drop")

dx_order <- map_dx %>%
  group_by(dx) %>%
  summarize(m = median(median_min), .groups = "drop") %>%
  arrange(m) %>%
  pull(dx)

map_dx_sf <- tidyr::expand_grid(
    fips = us$fips,
    year = sort(unique(map_dx$year)),
    dx   = dx_order
  ) %>%
  left_join(map_dx, by = c("fips" = "geography", "year", "dx")) %>%
  left_join(us, by = "fips") %>%
  mutate(
    year_lab = ifelse(as.character(year) %in% names(yr_lab),
                      yr_lab[as.character(year)], as.character(year)),
    dx = factor(dx, levels = dx_order)
  ) %>%
  st_as_sf()

# One shared scale so rows are comparable. It is capped at 900 minutes because a
# handful of state-years (mostly psychosis and behavioral) run to ~2,600 and an
# uncapped scale would flatten every other row to near-white; values above the cap
# are drawn at the cap.
ggplot(map_dx_sf) +
  geom_sf(aes(fill = median_min), colour = surf, linewidth = 0.1) +
  facet_grid(dx ~ year_lab, switch = "y") +
  scale_fill_gradientn(
    colours = seq_blue, na.value = grid_c, labels = label_number(),
    limits = c(150, 900), oob = scales::squish,
    guide = guide_colourbar(barheight = 10, barwidth = 0.8, ticks = FALSE)
  ) +
  labs(
    title = "Median ED stay by state, year and mental health diagnosis",
    subtitle = paste(
      "All sexes, monthly medians averaged within each year.",
      "Rows share one colour scale, so they are comparable to each other",
      sep = "\n"
    ),
    fill = "Median\nminutes",
    caption = paste(
      "Grey = no admissible value: suppressed at source (10 or fewer encounters), or a state absent from the export.",
      "Eating disorders is mappable in only 46% of state-years and behavioral in 90%; the other five are near-complete.",
      "Scale capped at 900 minutes; a few psychosis and behavioral state-years exceed it and are drawn at the cap.",
      sep = "\n"
    )
  ) +
  theme_map() +
  theme(
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    strip.placement   = "outside",
    panel.spacing     = unit(2, "pt")
  )
```

Rows are ordered by overall median, shortest at the top. Three things the grid shows that a
single combined map could not:

- **The geography is not the same for every diagnosis.** Psychosis and behavioral run long
  in a different set of states than substance use does, so any single mental health map
  averages away real variation.
- **Psychosis is both the longest row and the widest**, with far more spread between states
  than the other diagnoses — it would dominate any combined measure.
- **Eating disorders is mostly grey**, and behavioral is patchy in the sparser states. Those
  two rows should not be read as geographic patterns; they are closer to a map of where
  Epic's footprint is large enough to clear the suppression threshold.

```{r map-dx-table}
map_dx %>%
  filter(year == 2025) %>%
  select(geography_name, dx, median_min) %>%
  mutate(dx = factor(dx, levels = dx_order)) %>%
  pivot_wider(names_from = dx, values_from = median_min) %>%
  mutate(across(where(is.numeric), ~ round(.x))) %>%
  arrange(geography_name) %>%
  rename(State = geography_name)
```

The table gives 2025, the most recent complete year, so every mapped value is readable
without the colour scale. Blank cells are suppressed.

